Disentangling linkages between satellite derived forest structure and productivity across a range of forest types and ecosystems.

### 2023-04-18 scripts/4_vegan.qmd
added strata instead of just bec

```{r packages}
library("vegan")
```

@@

# normalization from grace et al 2016
# first does mylog (does natural logarithm. if 0s present, does log(one + x))
# then scales to z scores
# results it more normal looking data
normal <- function(data) {
  data %>%
    mylog() %>%
    scale() %>%
    as.numeric()
}
```

@@

@@

@@

summary(rda_out)$sites %>% 
  as.data.frame() %>% 
  select(RDA1) %>% 
  bind_cols(to_rda) %>%
  select(-bec, -category) %>% 
  cor() %>%
  as.data.frame() %>%
  select(RDA1) %>%
  arrange(desc(abs(RDA1)))

@@

varpart_out %>% 
  plot(Xnames = c("Derived", "Modelled"))

@@

bec_nested <- to_nest %>%
  group_by(bec) %>%
  nest() %>%
  ungroup()
```

@@

```{r rda-var-bec}
outs <- bec_nested %>%
  mutate(rda = map(data, my_rda_fun),
         var = map(data, my_var_fun)) 
```

@@

fracts_out <- outs %>%
  mutate(indfracts = map(var, .f = function(x) {
    x$part$indfract %>%
      mutate(type = c("extracted", "modelled", "overlap", "residual")) %>%
      select(type, Adj.R.squared) %>%
      pivot_wider(names_from = type, values_from = Adj.R.squared)
    }
    )) %>%
  unnest(indfracts) %>%
  arrange(desc(residual)) %>%
  select(bec, extracted, modelled, overlap) %>%
  pivot_longer(c(extracted, modelled, overlap)) %>%
  mutate(value = ifelse(value > 0, value, 0)) %>%
  pivot_wider()
```

```{r raincloud-plot}
library(ggdist)
fracts_out %>%
  left_join(keys$bec, by = c("bec" = "zone")) %>%
  pivot_longer(extracted:overlap) %>%
  mutate(name = str_to_title(name)) %>%
  mutate(value = ifelse(value < 0, 0, value)) %>%
  ggplot(aes(x = value, y = name)) +
  stat_halfeye(
    # adjust bandwidth
    adjust = .5,
    # move to the right
    justification = -0.2,
    # remove the slub interval
    .width = 0,
    point_colour = NA
    ) +
  geom_boxplot(
    width = 0.12,
    # removing outliers
    outlier.color = NA,
    alpha = 0.5
  ) +
  stat_dots(
    # ploting on left side
    side = "left",
    # adjusting position
    justification = 1.1,
    # adjust grouping (binning) of observations
    binwidth = 0.0025,
    #position = "stack",
    aes(fill = group, col = group),
    alpha = 0.5
  ) +

@@

fracts_out %>%
  pivot_longer(extracted:overlap) %>%
  left_join(bec_clim, multiple = "all") %>%
  filter(var == "ppt") %>%
  ggplot(aes(x = mean, y = value)) +
  geom_point() +
  geom_smooth(method = "lm") + 
  facet_wrap(~name) 
```

@@

```{r cdf-anova}
outs %>% 
  filter(bec == "CDF") %>%
  select(rda) %>% 
  pull(rda) %>%
  .[[1]] %>%
  anova()
```

@@

```{r eig-anova}
outs %>%
  mutate(eig = map(rda, .f = function(rda) {
    rda$CCA$eig
  })) %>%
  unnest_wider(eig) %>%
  mutate(prop21 = RDA1/RDA2) %>%
  arrange(desc(prop21)) %>%
  left_join(anova_a2) %>%
  select(bec, RDA1:sig)

@@

      anova_out$`Pr(>F)` %>%
        as.data.frame() %>%
        rename(anova_p = 1) %>%
        drop_na() %>%
        mutate(anova_sig = cut(
          anova_p,
          breaks = c(0, 0.001, 0.01, 0.05, .10, 1),
          labels = c("***", "**", "*", ".", "NS")))  %>%
        mutate(axis = paste0("RDA", row_number())) %>%
        relocate(axis)
  }))

anova_sig_axes <- anova_outs %>%
  select(bec, anova) %>%
  unnest(cols = anova) %>%
  filter(anova_sig != "NS")
```

@@

```{r}
out_loadings %>% 
  mutate(axis_sig = cut(
          axis_p,
          breaks = c(0, 0.001, 0.01, 0.05, .10, 1),
          labels = c("***", "**", "*", ".", "NS"),
          right = F)) %>%
  inner_join(anova_sig_axes) %>%
  #filter(axis_sig == "NS") %>%
  arrange(axis_loading) %>%
  count(var)

@@

need to get significant axes from the anova and then filter this

>>>>

```{r packages}
library("vegan")
library("tidyverse")
library("cols4all")
library(terra)
library(sf)
source(here::here("scripts", "get_keys.R"))
```

@@

# normalization from grace et al 2016
# first does mylog (does natural logarithm. if 0s present, does log(one + x))
# then scales to z scores
# results it more normal looking data
normal <- function(data) {
  data %>%
    mylog() %>%
    scale() %>%
    as.numeric()
}

```{r theme}
my_theme <- theme_bw() +
  theme(panel.grid = element_blank())

theme_set(my_theme)
```

@@

to_rda <- bec_normal %>% 
  group_by(category) %>% 
  slice_sample(n = 3000)  %>% 
  filter(category == "Mixed Wood") %>% 
  ungroup()

@@

rda_sum <- summary(rda_out)

var_exp <- rda_sum$constr.chi / rda_sum$tot.chi

rda_eig <- rda_out$CCA$eig %>%
  as.data.frame() %>%
  t() %>%
  as.data.frame() %>%
  tibble() %>%
  pivot_longer(cols = everything()) %>%
  mutate(prop = value / sum(value),
         var_exp = prop * var_exp)

@@

summary(rda_out)$sites %>% 
  as.data.frame() %>% 
  select(RDA1:RDA3) %>% 
  bind_cols(to_rda) %>%
  select(-bec, -category) %>% 
  cor() %>%
  as.data.frame() %>%
  select(RDA1:RDA3) %>%
  arrange(desc(abs(RDA1))) %>%
  filter(!(startsWith(row.names(.), "RDA")))

@@

var_out %>% 
  plot(Xnames = c("Derived", "Modelled"))

@@

bec_nested <- to_nest %>%
  group_by(bec) %>%
  nest() %>%
  ungroup() %>%
  rename(strata = bec)

class_nested <- to_nest %>%
  group_by(category) %>%
  slice_sample(n = 3000) %>%
  nest() %>% 
  ungroup() %>%
  rename(strata = category)

strata_nested <- bind_rows(bec_nested, class_nested)
```

@@

```{r rda-var-bec}
outs <- strata_nested %>%
  mutate(rda = map(data, my_rda_fun, .progress = "RDA"),
         var = map(data, my_var_fun, .progress = "VAR")) 
```

@@

fracts_out <- outs %>%
  mutate(indfracts = map(var, .f = function(x) {
    x$part$indfract %>%
      mutate(type = c("extracted", "modelled", "overlap", "residual")) %>%
      select(type, Adj.R.squared) %>%
      pivot_wider(names_from = type, values_from = Adj.R.squared)
    }
    )) %>%
  unnest(indfracts) %>%
  arrange(desc(residual)) %>%
  select(strata, extracted, modelled, overlap) %>%
  pivot_longer(c(extracted, modelled, overlap)) %>%
  mutate(value = ifelse(value > 0, value, 0)) %>%
  pivot_wider()

fracts_out %>% 
  mutate(total = extracted + modelled + overlap) %>%
  arrange(strata %in% (bec_normal %>%
           pull(category) %>%
           unique()),
          desc(total))
```

```{r raincloud-plot}
library(ggdist)
fracts_out %>%
  #left_join(keys$bec, by = c("bec" = "zone")) %>%
  pivot_longer(extracted:overlap) %>%
  mutate(name = str_to_title(name)) %>%
  mutate(value = ifelse(value < 0, 0, value)) %>%
  ggplot(aes(x = value, y = name)) +
  stat_halfeye(
    # adjust bandwidth
    adjust = .5,
    # move to the right
    justification = -0.2,
    # remove the slub interval
    .width = 0,
    point_colour = NA
    ) +
  geom_boxplot(
    width = 0.12,
    # removing outliers
    outlier.color = NA,
    alpha = 0.5
  ) +
  stat_dots(
    # ploting on left side
    side = "left",
    # adjusting position
    justification = 1.1,
    # adjust grouping (binning) of observations
    binwidth = 0.0025,
    #position = "stack",
    # aes(fill = group, col = group),
    alpha = 0.5
  ) +

@@

fracts_out %>%
  pivot_longer(extracted:overlap) %>%
  left_join(bec_clim, multiple = "all") %>%
  filter(var == "ppt") %>%
  ggplot(aes(x = mean, y = value)) +
  geom_point() +
  geom_smooth(method = "lm") + 
  facet_wrap(~name) +
  theme_bw() +
  theme(panel.grid = element_blank())
```

@@

```{r cdf-anova}
outs %>% 
  filter(bec == "IDF") %>%
  select(rda) %>% 
  pull(rda) %>%
  .[[1]] %>%
  anova(by = "axis")
```

@@

```{r eig-anova}
eig_outs <- outs %>%
  mutate(eig = map(
    rda,
    .f = function(rda) {
      rda_sum <- summary(rda)

      var_exp <- rda_sum$constr.chi / rda_sum$tot.chi

      rda_eig <- rda$CCA$eig %>%
        as.data.frame() %>%
        t() %>%
        as.data.frame() %>%
        tibble() %>%
        pivot_longer(cols = everything(), names_to = "axis",
                     values_to = "eigenvalue") %>%
        mutate(prop = eigenvalue / sum(eigenvalue),
               var_exp = prop * var_exp)
    }
  )) %>%
  unnest(eig) %>%
  select(strata, axis:var_exp)

@@

      anova_out$`Pr(>F)` %>%
        as.data.frame() %>%
        rename(anova_p = 1) %>%
        drop_na() %>%
        mutate(anova_sig = cut(
          anova_p,
          breaks = c(0, 0.001, 0.01, 0.05, .10, 1),
          labels = c("***", "**", "*", ".", "NS")))  %>%
        mutate(axis = paste0("RDA", row_number())) %>%
        relocate(axis)
  }, .progress = "anovas"))

anova_sig_axes <- anova_outs %>%
  select(bec, anova) %>%
  unnest(cols = anova) # %>%
  # filter(anova_sig != "NS")

axis_varexp <- anova_sig_axes %>%
  inner_join(eig_outs)

axis_varexp %>%
  left_join(keys$bec, by = c("bec" = "zone")) %>%
  ggplot(aes(x = axis, y = var_exp, col = axis)) +
  geom_jitter(height = 0, width = 0.05) +
  theme(legend.position = "bottom") +
  labs(x = NULL,
       y = "% of Variance Explained",
       col = NULL) +
  scale_y_continuous(labels = scales::label_percent())
# no secondary axis in alpine & coast.

axis_varexp %>%
  group_by(bec) %>%
  # filter(n() == 2) %>%
  select(bec, axis, var_exp) %>% 
  pivot_wider(names_from = axis,
              values_from = var_exp) %>%
  mutate(ratio = RDA1 / RDA2)

# smaller ratios present when the RDA1 doesn't explain much. when variance explained
# is low, the ratio is low. this is because there isn't much variance explained
```

@@

```{r}
sig_loads <- out_loadings %>% 
  mutate(loading_sig = cut(
          axis_p,
          breaks = c(0, 0.001, 0.01, 0.05, .10, 1),
          labels = c("***", "**", "*", ".", "NS"),
          right = F)) %>%
  inner_join(anova_sig_axes) %>%
  #filter(axis_sig == "NS") %>%
  arrange(desc(abs(axis_loading))) %>%
  filter(!(var %in% c("percentage_first_returns_above_mean", 
                    "elev_mean", 
                    "elev_stddev"))) %>%
  drop_na()# removes ima mindhi which has no variance

sig_loads %>%
  filter(endsWith(var, "DHI"),
         axis == "RDA2") %>%
  arrange(bec) %>%
  select(bec, var, axis_loading) %>%
  pivot_wider(names_from = var,
              values_from = axis_loading)
# when a second axis is present, it has a weak loading on the cumulative DHI
# while the majoirty of the variability is contained within all three DHIs together, the second axis, when present, is more linked to the minmimum and variation DHI.

sig_loads %>%
  select(bec, var, axis, axis_loading, loading_sig) %>%
  filter(endsWith(var, "DHI")) %>%
  ggplot(aes(x = var, y = axis_loading, group = bec)) +
  geom_point(aes(col = bec)) +
  geom_line(aes(col = bec), alpha = 0.5, lty = "dashed") +
  geom_hline(yintercept = 0, col = "red", alpha = 0.2) +
  facet_wrap(~axis) +
  theme_bw() +
  theme(panel.grid = element_blank()) +
  labs(x = NULL,
       y = "Axis Loading")

sig_loads %>%
  select(bec, var, axis, axis_loading) %>%
  filter(!endsWith(var, "DHI")) %>%
  left_join(keys$continuous, by = c("var" = "variable")) %>%
  ggplot(aes(x = var_long, y = abs(axis_loading), group = bec)) +
  geom_point(aes(col = bec)) +
  geom_line(aes(col = bec), alpha = 0.5, lty = "dashed") +
  geom_hline(yintercept = 0, col = "red", alpha = 0.2) +
  facet_wrap(~axis) +
  theme(axis.text.x = element_text(angle = 45,
                                   hjust = 1)) +
  labs(x = NULL,
       y = "Axis Loading (Absolute Value)")

sig_loads %>%
  filter(!endsWith(var, "DHI"),
         loading_sig != "NS") %>%
  group_by(bec, axis) %>%
  mutate(loading_rank = data.table::frankv(abs(axis_loading), 
                                           order = -1)) %>%
  ggplot(aes(x = fct_reorder(var, loading_rank),
             y = loading_rank, col = bec)) +
  geom_jitter(height = 0) +
  facet_wrap(~axis) +
  theme(axis.text.x = element_text(angle = 45, 
                                   hjust = 1)) +
  labs(x = NULL,
       y = "Axis Loading Rank")

@@

```{r jenkies}
library(rgeoda)
sig_loads_f <- sig_loads %>%
  filter(!endsWith(var, "DHI"),
         loading_sig != "NS")

jenkies <- sig_loads_f %>%
  mutate(axis_loading = abs(axis_loading)) %>%
  group_by(bec, axis) %>%
  select(bec, axis, axis_loading) %>%
  nest() %>%
  mutate(breaks = map(data, natural_breaks, k = 2)) %>%
  unnest(breaks) %>%
  select(-data)

sig_loads_f %>%
  ggplot(aes(x = bec, y = abs(axis_loading))) +
  facet_wrap(~axis) +
  geom_point(data = jenkies, aes(y = breaks), col = "red", size = 3) +
  geom_point(aes(shape = var))

sig_loads_f %>%
  left_join(jenkies) %>%
  mutate(strong = abs(axis_loading) > breaks) %>%
  ggplot(aes(x = bec, y = var)) +
  geom_tile(aes(fill = strong, alpha = abs(axis_loading)))
  geom_point(aes(col = strong))
  left_join(keys$bec, by = c("bec" = "zone")) %>%
  group_by(var) %>%
  summarize(per_strong = sum(strong) / n())

# We identified large contributors to each axis using a thresholding process. The threshold for each BEC zone and significant RDA axis was determined by using the Jenks natural breaks optimization method with k = 2 on significant loadings. We then summarized the most common large contributors across BEC zones and our BEC groupings.
```

```{r cutoffs}
for_cutoff_plot <- sig_loads_f %>%
  select(bec, axis, var, axis_loading) %>%
  crossing(cutoff = seq(0, 1, 0.01)) %>%
  mutate(over_cut = abs(axis_loading) > cutoff) %>%
  group_by(var, cutoff) %>%
  summarize(prop_over_cut = sum(over_cut) / n())

for_cutoff_plot %>%
  left_join(keys$continuous, by = c("var" = "variable")) %>%
  ggplot(aes(x = cutoff, y = prop_over_cut, col = var_long)) +
  geom_line() +
  theme(legend.position = c(0.8, 0.6)) +
  labs(x = "Cutoff Value",
       y = "Proportion of loadings above cutoff",
       col = "Variable") +
  scale_y_continuous(labels = scales::label_percent()) +
  scale_color_discrete_c4a_cat(palette = "met.tsimshian") +
  geom_vline(xintercept = c(0.2, 0.25, 0.3), col = "red", lty = "dashed")
```

calculate the actual variance explained

### 2023-04-18 scripts/manuscript/manuscript.qmd
added strata instead of just bec

# include all libraries required by the R chunks here
library(lavaan)
library(sgsR)
library(terra)
library(sf)

@@

The overall goal of this paper is to characterize patterns of forest
ecosystem structure and function and their interrelationships across a
wide range of ecosystems encompassing significant environmental
gradients. To do so, we synthesize data from moderate-scale
remote-sensing derived metrics of structure, represented as both simple
ALS metrics of canopy height and cover, as well as modeled forest
structure variables including volume, and above ground biomass, with a
remote sensing derived index on ecosystem function. Our first question
is to examine how ecosystem structure and function vary across a large
environmental gradient and then compare the simple and modeled
representations of forest structure to different levels of ecosystem
function. This question is important as it provides insights to the EBV
community around redundancy in remote sensing metrics when describing
structure and function on ecosystem and proposes a method to examine
potential overlap when generating remote sensing EBV's.

@@

British Columbia is the westernmost province of Canada, and is home to a
variety of terrestrial ecosystems [@pojar1987]. Approximately 64% of the
province is forested [@bcministryofforests2003]. There is a large amount
of ecosystem variation in the province, with large climate and
topographic gradients. The Biogeoclimatic Ecosystem Classification (BEC)
system identifies 16 zones based on the dominant tree species and the
ecosystems general climate. These zones can be further split into
subzones, variants, and phases based on microclimate, precipitation, and
topography [@pojar1987].

@@

We conducted model based sampling across the fifteen forest dominated
ecosystems found within British Columbia \[BC figure reference, I have
not made this figure yet\]. Samples were randomly selected within BEC
zones of interest alongside multiple criteria. Each sampled pixel had to
have a forested land cover class (coniferous, deciduous, mixed-wood, or
wetland-treed), and be surrounded by the same land cover class.
Additionally, each pixel had to have a coefficient of variation less
than 0.5 in the Lorey's height and canopy cover forest structural
metrics. A maximum of 3000 samples were sampled in each BEC zone. To
meet the normality assumptions of structural equation modelling, all
variables were natural-log transformed and standardized, as per
@grace2016. Variables containing zeros were natural-log plus one
transformed.

@@

### Path Analysis

To determine the relationships between simple lidar derived forest
structural attributes, complex/derived forest structural attributes, and
forest productivity, we used path analysis to analyze two causal models
(@fig-pathdag). To determine primary drivers of the three DHIs, we will
summarize the predictors across ecosystems by counting the strongest
predictor in each BEC zone. This will determine if the primary driver
for each facet of the DHI is simple or complex, and allow us to assess
ecosystem differences.

```{r}
#| label: fig-pathdag
#| fig-cap: Proposed path diagrams.
#| warning: false
#| echo: false

knitr::include_graphics(here::here("outputs", "path_dags.png"))
```

### Exploratory Structural Equation Modelling

Secondly, we will use exploratory structural equation modelling
[@marsh2020; @asparouhov2009] to identify latent forest structural
variables within the data, and use these latent variables to predict the
dynamic habitat indices. Exploratory structural equation modelling is a
combination of exploratory factor analysis (EFA) and structural equation
modelling, which relaxes the strict requirement of zero cross-loadings
in confirmatory factor analysis and allows for less strict measurement
models to be used. ESEM is used when it is known that there is a latent
structure to the data, but the specific indicator variables have not yet
been determined.

One advantage of ESEM is that it can create varying numbers of latent
variables. For our analysis, we first ran exploratory factor analysis on
all forest structural attributes both globally and by BEC zone with
between 1-4 potential latent variables. We chose the most parsimonious
EFA model with the lowest AIC scores for each ecosystem, leading to
varying numbers of latent variables. We then determined the anchoring
indicator for each latent variable. The anchors were calculated by
choosing the indicator variable with the largest difference between the
maximum value in a given loading compared to to said indicators loadings
in all other latent variables. Each anchor variable was then assigned to
a named group (Canopy Cover, Height and Biomass, and Structural
Complexity).

Following the EFA, we used structural equation modelling, with latent
variables loadings determined by the EFA. We also filtered the loadings
to be greater than or equal to 0.5, allowing us to examine the number
and composition of latent variables found in each ecosystem. Following
this, we predicted the DHIs in a single SEM with covariances between the
DHIs. If a forest structural attribute did not end up in a latent
variable, it was included in the structural equation model as a DHI
predictor, without a latent variable as a mediator.

@@

The processing for this paper was done in R version 4.2.2 [@R-base]. We
used *sgsR* [@goodbody2023] for sampling, *terra* [@R-terra] for raster
analyses, *sf* for vector analyses [@R-sf], and *lavaan* [@lavaan2012]
for path analysis and structural equation modelling.

@@

## Path Analysis

```{r}
#| label: fig-pathbar-prop
#| fig-cap: Proportion of strongest signficant forest structural predictors of yearly productivity metrics in models based on canopy cover vs canopy height.
#| warning: false
#| echo: false

knitr::include_graphics(here::here("outputs", "path_bars.png"))
```

```{r}
#| label: fig-pathbar-pred
#| fig-cap: Bar plot of strongest predictors in path analysis for each BEC zone and predicted DHI variable.
#| warning: false
#| echo: false

knitr::include_graphics(here::here("outputs", "strongest_predictor_bar.png"))
```

```{r}
#| label: fig-pathmap
#| fig-cap: Map of strongest structural attribute predictor strength on yearly productivity metrics by BEC zone for the two proposed models. If there are no significant predictors, shown in dark grey
#| warning: false
#| echo: false

knitr::include_graphics(here::here("outputs", "path_map_edit.png"))
```

## Exploratory Structural Equation Modelling

I also want to run exploratory SEM with and without the DHI's to see
what grouping it goes in. MinDHI has been breaking because some zones
only have 0 as mindhi (makes sense, they are in the alpine).

```{r}
#| label: fig-esem-latent
#| fig-cap: Groups of latent variables with loadings greater than or equal to 0.5 for each BEC zone. Loadings were determined using exploratory factor analysis with up to four latent variables, and selecting the number of possible latent variables with the lowest AIC. All forest structure variables were potentially included as indicators in each latent variable.
#| warning: false
#| echo: false

knitr::include_graphics(here::here("outputs", "latent_groups.png"))
```

Exploratory factor analysis shows three groups of latent variables
(@fig-esem-latent). We assigned the groups to latent variables
indicating canopy cover / biomass metrics, canopy height / biomass
metrics, and structural complexity metrics. Indicators to latent
variable groups by examining anchors for each latent variable. The
anchors were calculated by choosing the indicator variable with the
largest difference between the maximum value in a given loading compared
to to said indicators loadings in all other latent variables.

Both the Ponderosa Pine (PP) and Sub-boreal Pine -- Spruce zones have
one latent variable. Coastal Western Hemlock (CWH), Engelmann Spurce --
Subalpine Fir (ESSF), and Montane Spruce (MS) are the only zones to have
structural complexity based latent variables. CWH is the only zone to
include elevation standard deviation in the structural complexity latent
variable.

```{r}
#| label: fig-esem-latent-dhi
#| fig-cap: Groups of latent variables with loadings greater than or equal to 0.5 for each BEC zone, including the DHIs. Loadings were determined using exploratory factor analysis with up to five latent variables, and selecting the number of possible latent variables with the lowest AIC. All forest structure variables were potentially included as indicators in each latent variable.
#| warning: false
#| echo: false

knitr::include_graphics(here::here("outputs", "latent_dhis.png"))
```

Exploratory factor analysis shows three groups of latent variables. We
assigned these groups to attributes based on the indicator variables in
each latent variable, indicating canopy cover, height and biomass, and
structural complexity. When including the DHIs in this EFA, it is rare
that they are included with latent variables with large loadings. The
cumulative DHI is an indicator for the canopy cover latent varible in
the IDF (Interior Douglas-fir) and MS (Montane Spruce) BEC zones. The
DHIs are not included in any other latent variable, indicating that they
compose different information than the forest structural variables.

```{r}
#| label: fig-latent-boxplots
#| fig-cap: Boxplots of latent variable loadings across BEC zones as determined by exploratory factor analysis. Red dashed lines show latent variable indicator cutoff before being used as starting values for the SEM models.
#| warning: false
#| echo: false

knitr::include_graphics(here::here("outputs", "latent_boxplot.png"))
```

@fig-latent-boxplots shows boxplots of latent variable loadings from
forest structure indicator variables. A maximum of three valid latent
variables were identified across the sixteen BEC zones. We assigned
indicators variables to latent variable groups by examining the
anchoring variables for each latent variable. The anchoring variables
were calculated by choosing the indicator variable with the largest
difference between the maximum value in a given loading compared to said
indicators loadings in all other latent variables. The only DHI variable
that was included in the latent variables after subsetting the latent
variables to only include strong loadings was the Cumulative DHI in the
canopy cover latent variable.

I need to actually make the DAGs and plot the parameters, but I haven't
gotten there yet. I'll plot the global one to show what it looks like,
then aggregate it in some way - probably similar to the path analysis. I
also may include the variable partitioning that Meg talked about in the
email if I can get it working.

>>>>

# include all libraries required by the R chunks here
library(vegan)
library(sgsR)
library(terra)
library(sf)

@@

The overall goal of this paper is to characterize patterns of forest
ecosystem structure and function and their interrelationships across a
wide range of ecosystems encompassing significant environmental
gradients. To do so, we synthesize data from moderate-scale
remote-sensing derived metrics of structure, represented as both simple
ALS metrics of canopy height and cover, as well as modeled forest
structure variables including volume, and above ground biomass, with a
remote sensing derived index on ecosystem function. Our first question
is to examine how ecosystem structure and function vary across a large
environmental gradient and then compare the simple and modeled
representations of forest structure to different levels of ecosystem
function. This question is important as it provides insights to the EBV
community around redundancy in remote sensing metrics when describing
the structure and function of ecosystems and proposes a method to
examine potential overlap when generating remote sensing EBV's.

@@

British Columbia is the westernmost province of Canada, and is home to a
variety of terrestrial ecosystems [@pojar1987]. Approximately 64% of the
province is forested [@bcministryofforests2003]. There is a large amount
of ecosystem variation in the province, with large climate and
topographic gradients. The Biogeoclimatic Ecosystem Classification (BEC)
system identifies 16 zones based on the dominant tree species and the
ecosystems general climate. These zones can be further split into
subzones, variants, and phases based on microclimate, precipitation, and
topography [@pojar1987]. To examine trends across the large
environmental gradients, we group the BEC zones into five broad biomes,
specifically, the south interior, northern, montane, alpine, and coastal
groups similar to @hamann2006.

@@

We conducted model based sampling across the fifteen forest dominated
ecosystems found within British Columbia \[BC figure reference, I have
not made this figure yet\]. Samples were randomly selected within BEC
zones of interest alongside multiple criteria. Each sampled pixel had to
have a forested land cover class (coniferous, deciduous, mixed-wood, or
wetland-treed), and be surrounded by the same land cover class.
Additionally, each pixel had to have a coefficient of variation less
than 0.5 in the Lorey's height and canopy cover forest structural
metrics. A maximum of 3000 samples were sampled in each BEC zone. All
variables were natural-log transformed and standardized. Variables
containing zeros were natural-log plus one transformed.

@@

### Partial Redundancy Analysis and Variation Partitioning

Partial redundancy analysis and variation partitioning was used to
relate the primary and modeled forest structure variables to ecosystem
structure, to assess the inter-relationships on forest ecosystem
function across the broad environment range. Redundancy analysis has
widely been used in community ecology where environmental variables of
interest are compared to species composition, and when the effect of
space for example is assessed as a third covariate, in an approach known
as partial redundancy analysis. While RDA is common in the ecological
literature, this represents one of the first times this technique has
been applied to assess redundancy in proposed essential biodiversity
variables. Redundancy analysis utilizes multiple regression in that it
models the effect of an explanatory matrix on a response matrix; with
the key difference being the use of a response matrix rather than a
single response variable. The first step is a multiple regression where
each of the variables in the response and explanatory matrices are
regressed resulting in a matrix of fitted values. PCA is then used in
the second step to reduce dimensionality through eigenvectors and to
allow the strength of relationships between assessed, taking into
account the number of explanatory variables. Partial Redundancy Analysis
(PRA) allows the use of additional explanatory variables, or
co-variates. This is useful, as it allow the assessment of the effect of
one set of variables on another while taking into account variations due
to other third set that are not the focus of the study. The output is
similar to standard RDA except it is also possible to examine how much
variation is explained in the primary variables of interest, and the
co-variates alone, and then when combined together. This variance
partitioning is then shown typically through a Venn diagram, in which
the percentage of explained variance by each explanatory data set is
reported. We use ANOVAs to test which canonical axes are significant,
and eigenvalues to determine the proprotion of variance explained by
each axis.

@@

The processing for this paper was done in R version 4.2.2 [@R-base]. We
used *sgsR* [@goodbody2023] for sampling, *terra* [@R-terra] for raster
analyses, *sf* for vector analyses [@R-sf], and *vegan* [@R-vegan] for
redundancy analysis and variation partitioning.

@@
